#Load Required Packages
```{r}
library(edgeR)
library(tximport)
library(svglite)
library(dplyr)
library(ggplot2)
library(ggvenn)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)
library(pheatmap)
```

#Set file path and file names
```{r}
file_dir <- "/project/pi_cpak_umass_edu/Jay_Bulk_RNAseq/RSEM/RSEM_counts"
files <- list.files(file_dir, pattern = "\\.genes\\.results$", full.names = TRUE)
```

#Rename files to reflect Sample ID + Treatment
```{r}
rename_table <- tribble(
  ~old,              ~new,
  "25298R-01-01v1", "3320a-1_TTX",
  "25298R-01-02v1", "3320a-2_TTX",
  "25298R-01-03v1", "3320a-3_TTX",
  "25298R-01-04v1", "3320a-1",
  "25298R-01-05v1", "3320a-2",
  "25298R-01-06v1", "3320a-3",
  "25298R-01-07v1", "7880-1_TTX",
  "25298R-01-08v1", "7880-2_TTX",
  "25298R-01-09v1", "7880-3_TTX",
  "25298R-01-10v1", "7880-1",
  "25298R-01-11v1", "7880-2",
  "25298R-01-12v1", "7880-3",
  "25298R-01-13v1", "C3141a-1_TTX",
  "25298R-01-14v1", "C3141a-2_TTX",
  "25298R-01-15v1", "C3141a-3_TTX",
  "25298R-01-16v1", "C3141a-1",
  "25298R-01-17v1", "C3141a-2",
  "25298R-01-18v1", "C3141a-3"
)

old_ids <- gsub("\\.genes\\.results$", "", basename(files))
sample_names <- rename_table$new[match(old_ids, rename_table$old)]

names(files) <- sample_names
```

#Drop TTX treated samples 
```{r}
drop_samples <- c("C3141a-1_TTX","C3141a-2_TTX","C3141a-3_TTX",
                  "3320a-1_TTX","3320a-2_TTX","3320a-3_TTX",
                  "7880-1_TTX","7880-2_TTX","7880-3_TTX")

files <- files[!names(files) %in% drop_samples]
```

#Load in files using TxImport + Add cleaned up gene names to columns
```{r}
txi <- tximport(files, type = "rsem", txIn = FALSE, txOut = FALSE)

clean_gene_names <- function(ids) {
  gsub("^[^_]*_(.*)$", "\\1", ids)
}

clean_ids <- clean_gene_names(rownames(txi$counts))
rownames(txi$counts)  <- clean_ids
rownames(txi$abundance) <- clean_ids
rownames(txi$length)  <- clean_ids
```

#Add metadata for conditions 
```{r}
metadata <- tibble(
  sample = colnames(txi$counts)
) %>%
  mutate(
    condition = case_when(
      grepl("^C3141a", sample) ~ "Control",
      grepl("^7880",   sample) ~ "ASD",
      grepl("^3320a",  sample) ~ "SCZ",
      TRUE ~ NA_character_)
  )
```

#Turn metadata to a dataframe
```{r}
metadata_df <- as.data.frame(metadata)
rownames(metadata_df) <- metadata_df$sample

metadata_df$condition <- factor(metadata_df$condition,
                                levels = c("Control", "ASD", "SCZ"))

table(metadata_df$condition)
```

#Set up DGEList for EdgeR + Check counts per sample
```{r}
dge <- DGEList(counts = round(txi$counts), samples = metadata_df)

colSums(dge$counts)
```

#Create EdgeR design matrix for conditions 
```{r}
design <- model.matrix(~ condition, data = metadata_df)
colnames(design) <- make.names(colnames(design))
design
```

#Filter counts by expression using the design matrix + normalize library sizes 
```{r}
keep <- filterByExpr(dge, design)
dge <- dge[keep, , keep.lib.sizes=FALSE]
dge <- normLibSizes(dge)
```

#Estimate Dispersion
```{r}
dge <- estimateDisp(dge, design)
```

#Save DGE_List after filtering + normalizing 
```{r}
saveRDS(dge, "DGE_List.rds")
```

#Plot samples in MDS
```{r}
#svglite("MDS_plot.svg", width = 8, height = 6)
plotMDS(dge, col = as.numeric(metadata_df$condition))
#dev.off()
```
#Use glmQL for fit
```{r}
fit <- glmQLFit(dge, design)
```

#Condition Analysis 
```{r}
contrasts <- makeContrasts(
  ControlvASD = conditionASD,
  ControlvSCZ = conditionSCZ,
  levels = design
)
```

#Find ASD/SCZvControl DEGs
```{r}
ASDvControlDEG <- glmQLFTest(fit, contrast=contrasts[,"ControlvASD"])
SCZvControlDEG <- glmQLFTest(fit, contrast=contrasts[,"ControlvSCZ"])
```

#Make table of DEG Results
```{r}
res_ASD <- topTags(ASDvControlDEG, n = Inf)$table
res_SCZ <- topTags(SCZvControlDEG, n = Inf)$table
```

#Filter results for FDR <= 0.05
```{r}
res_ASD_sig <- res_ASD[res_ASD$FDR <= 0.05, ]
res_SCZ_sig <- res_SCZ[res_SCZ$FDR <= 0.05, ]
```

#Turn results into a table with the gene names as the columns
```{r}
res_ASD_sig <- as.data.frame(res_ASD_sig) %>% 
  rownames_to_column("gene")
res_ASD <- as.data.frame(res_ASD) %>% 
  rownames_to_column("gene")

res_SCZ_sig <- as.data.frame(res_SCZ_sig) %>% 
  rownames_to_column("gene")
res_SCZ <- as.data.frame(res_SCZ) %>% 
  rownames_to_column("gene")
```

#Print output DEGs as .csv files (raw + FDR filtered)
```{r}
write_csv(as.data.frame(res_ASD_sig, rownames = "gene"), "ControlvASD_DEGs.csv")
write_csv(as.data.frame(res_SCZ_sig, rownames = "gene"), "ControlvSCZ_DEGs.csv")
write_csv(as.data.frame(res_ASD, rownames = "gene"), "ControlvASD_DEGsALL.csv")
write_csv(as.data.frame(res_SCZ, rownames = "gene"), "ControlvSCZ_DEGsALL.csv")
```

#Volcano Plots
#Load in DEG outputs
```{r}
res_ASD_sig <- read.csv("ControlvASD_DEGs.csv")
res_SCZ_sig <- read.csv("ControlvSCZ_DEGs.csv")
res_ASD <- read.csv("ControlvASD_DEGsALL.csv")
res_SCZ <- read.csv("ControlvSCZ_DEGsALL.csv")
```

#Create Volcano Plots for ASD DEGs using EnhancedVolcano
```{r}
#svglite("VolcanoPlot_ControlvASD_NoFC.svg", width = 14, height = 10)
EnhancedVolcano(
  res_ASD,
  lab = res_ASD$gene,
  FCcutoff = 0,
  x = "logFC",
  y = "PValue",
  title = "ASD vs Control"
)
#dev.off()
```
#Create Volcano Plots for SCZ DEGs using EnhancedVolcano
```{r}
#svglite("VolcanoPlot_ControlvSCZ_NoFC.svg", width = 14, height = 10)
EnhancedVolcano(
  res_SCZ,
  lab = res_SCZ$gene,
  FCcutoff = 0,
  x = "logFC",
  y = "PValue",
  title = "SCZ vs Control"
)
#dev.off()
```

#Venn Diagram
#Create Venn Diagram of ASD/SCZvControl DEGs using ggvenn
```{r}
asd_genes <- res_ASD_sig$gene
scz_genes <- res_SCZ_sig$gene

p <- ggvenn(list(ASD = asd_genes, SCZ = scz_genes),
       fill_color = c("pink", "grey"),
       stroke_size = 0.5,
       set_name_size = 5,
       text_size = 4)

ggsave("DEG_VennDiagram.svg", plot = p, width = 8, height = 6)
```

#ClusterProfiler
#Load in DEG results 
```{r}
res_ASD_sig <- read.csv("ControlvASD_DEGs.csv")
res_SCZ_sig <- read.csv("ControlvSCZ_DEGs.csv")
All_res_ASD <- read.csv("ControlvASD_DEGsALL.csv")
All_res_SCZ <- read.csv("ControlvSCZ_DEGsALL.csv")
```

#Set gene name variable 
```{r}
asd_symbols <- res_ASD_sig$gene
scz_symbols <- res_SCZ_sig$gene
gene_symbols <- All_res_ASD$gene
```

#Convert from gene symbols to ENTREZID for ASD DEGs
```{r}
asd_genes <- bitr(
  asd_symbols,
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
)
```
#Add FC to genes differentially expressed for ASD
```{r}
ASD_FC <- res_ASD_sig$logFC[match(asd_genes$SYMBOL, res_ASD_sig$gene)]
names(ASD_FC) <- asd_genes$ENTREZID
```

#Convert from gene symbols to ENTREZID for SCZ DEGs
```{r}
scz_genes <- bitr(
  scz_symbols,
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
)
```
#Add FC to genes differentially expressed for SCZ
```{r}
SCZ_FC <- res_SCZ_sig$logFC[match(scz_genes$SYMBOL, res_SCZ_sig$gene)]
names(SCZ_FC) <- scz_genes$ENTREZID
```

#Convert from gene symbols to ENTREZID for all genes in analysis (Background)
```{r}
all_genes <- bitr(
  gene_symbols,
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
)
```

#Run enrichGO on ASD DEGs for BP 
```{r}
ASD_ego <- enrichGO(gene      = asd_genes$ENTREZID,
                universe      = all_genes$ENTREZID,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
```

#Run enrichGO on SCZ DEGs for BP
```{r}
SCZ_ego <- enrichGO(gene      = scz_genes$ENTREZID,
                universe      = all_genes$ENTREZID,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
```

#Plot bar plots of enrichGO results 
```{r}
svglite("ClusterProfilerBarPlot_ControlvASD.svg", width = 14, height = 10)
barplot(ASD_ego, showCategory = 15) +
  ggtitle("ASD – GO BP enrichment")
dev.off()

svglite("ClusterProfilerBarPlot_ControlvSCZ.svg", width = 14, height = 10)
barplot(SCZ_ego, showCategory = 15) +
  ggtitle("SCZ – GO BP enrichment")
dev.off()
```

#Plot dot plots of enrichGO results 
```{r}
svglite("ClusterProfilerDotPlot_ControlvASD.svg", width = 14, height = 10)
dotplot(ASD_ego, showCategory = 15) +
  ggtitle("ASD – GO BP enrichment")
dev.off()

svglite("ClusterProfilerDotPlot_ControlvSCZ.svg", width = 14, height = 10)
dotplot(SCZ_ego, showCategory = 15) +
  ggtitle("SCZ – GO BP enrichment")
dev.off()
```
#Plot network plots 
```{r}
#svglite("ClusterProfilerConnectionPlot_ControlvASD.svg", width = 14, height = 10)
p1 <- cnetplot(ASD_ego, showCategory = 8, foldChange = ASD_FC, layout = "kk", fc_threshold = 0.05) 
p1 <- p1 + scale_colour_gradient2(
    low       = "red",
    mid = "white",
    high      = "green",
    midpoint  = 0
  )
print(p1)
#dev.off()

#svglite("ClusterProfilerConnectionPlot_ControlvSCZ.svg", width = 14, height = 10)
p2 <- cnetplot(SCZ_ego, showCategory = 8, foldChange = SCZ_FC, layout = "kk", fc_threshold = 0.05)
p2 <- p2 + scale_colour_gradient2(
    low       = "red",
    mid = "white",
    high      = "green"
  )
print(p2)
#dev.off()
```

#Print Go Terms as a dataframe 
```{r}
asd_go_df <- as.data.frame(ASD_ego)
scz_go_df <- as.data.frame(SCZ_ego)
```

#Print ClusterProfiler results in a .csv file 
```{r}
write_csv(asd_go_df, "ClusterProfiler_GO_ASD.csv")
write_csv(scz_go_df, "ClusterProfiler_GO_SCZ.csv")
```

#Print ASD specific, SCZ specific, and shared DEGs
```{r}
ASD_Specific <- setdiff(asd_genes$ENTREZID, scz_genes$ENTREZID)
SCZ_Specific <- setdiff(scz_genes$ENTREZID, asd_genes$ENTREZID)
Shared <- intersect(asd_genes$ENTREZID, scz_genes$ENTREZID)
```

#Run enrichGO for BP of ASD specific genes 
```{r}
ASD_only_ego <- enrichGO(gene = ASD_Specific,
                universe      = all_genes$ENTREZID,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
```

#Run enrichGO for BP of SCZ specific genes
```{r}
SCZ_only_ego <- enrichGO(gene = SCZ_Specific,
                universe      = all_genes$ENTREZID,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
```

#Run enrichGO for BP of ASD/SCZ shared genes
```{r}
Shared_ego <- enrichGO(gene = Shared,
                universe      = all_genes$ENTREZID,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
```

#Plot dot plots of enrichGO results 
```{r}
svglite("GO_DotPlot_ASD_specific.svg", width = 14, height = 10)
dotplot(ASD_only_ego, showCategory = 15) +
  ggtitle("ASD-specific DEGs – GO BP enrichment")
dev.off()

svglite("GO_DotPlot_SCZ_specific.svg", width = 14, height = 10)
dotplot(SCZ_only_ego, showCategory = 15) +
  ggtitle("SCZ-specific DEGs – GO BP enrichment")
dev.off()

svglite("GO_DotPlot_Shared_ASD_SCZ.svg", width = 14, height = 10)
dotplot(Shared_ego, showCategory = 15) +
  ggtitle("Shared ASD/SCZ DEGs – GO BP enrichment")
dev.off()
```
#Create dataframe of GO terms 
```{r}
asd_only_go_df <- as.data.frame(ASD_only_ego)
scz_only_go_df <- as.data.frame(SCZ_only_ego)
shared_go_df <- as.data.frame(Shared_ego)
```

#Print ClusterProfiler results as .csv  
```{r}
write_csv(asd_only_go_df, "ClusterProfiler_GO_ASD_Only.csv")
write_csv(scz_only_go_df, "ClusterProfiler_GO_SCZ_Only.csv")
write_csv(shared_go_df, "ClusterProfiler_GO_Shared.csv")
```

#Shared Comparison 
#Load in FDR sig DEGs
```{r}
res_ASD_sig <- read.csv("ControlvASD_DEGs.csv")
res_SCZ_sig <- read.csv("ControlvSCZ_DEGs.csv")
```

#Print shared genes of ASD/SCZvControl DEGs 
```{r}
shared <- intersect(res_ASD_sig$gene, res_SCZ_sig$gene)
```

#Print ASD/SCZ genes that are in the shared DEGs
```{r}
asd_sub <- res_ASD_sig %>% filter(gene %in% shared)
scz_sub <- res_SCZ_sig %>% filter(gene %in% shared)
```

#Add FC to shared DEGs 
```{r}
merged <- inner_join(
  asd_sub %>% dplyr::select(gene, logFC_ASD = logFC),
  scz_sub %>% dplyr::select(gene, logFC_SCZ = logFC),
  by = "gene"
)
```

#Print delta FC of shared DEGs 
```{r}
merged <- merged %>%
  mutate(deltaFC = logFC_ASD - logFC_SCZ)

merged <- merged %>%
  dplyr::mutate(
    deltaFC = logFC_ASD - logFC_SCZ,
    abs_deltaFC = abs(deltaFC)
  )
```

#Print results of shared DEGs 
```{r}
write_csv(merged, "Shared_DEG_FC.csv")
```

#Look at top 50 by delta FC 
```{r}
top50 <- merged %>%
  dplyr::arrange(desc(abs_deltaFC)) %>%
  dplyr::slice(1:50)
```

#Create matrix of top 50 genes and print a heatmap using pheatmap
```{r}
fc_matrix <- top50 %>%
  dplyr::select(gene, logFC_ASD, logFC_SCZ) %>%
  tibble::column_to_rownames("gene")

pheatmap(
  fc_matrix,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  main = "Shared DEG Fold Changes"
)
```

#Categorize DEGs are Up in both/Down in both/Opposite directions + Print plot
```{r}
merged2 <- merged %>%
  mutate(
    category = case_when(
      logFC_ASD > 0 & logFC_SCZ > 0 ~ "Up in both",
      logFC_ASD < 0 & logFC_SCZ < 0 ~ "Down in both",
      TRUE ~ "Opposite directions"
    )
  )

N <- 50

lab_df <- merged2 %>%
  group_by(category) %>%
  slice_max(order_by = abs_deltaFC, n = N, with_ties = FALSE) %>%
  ungroup()

pal <- c(
  "Up in both" = "red",   
  "Down in both" = "blue",
  "Opposite directions" = "forestgreen"
)

ggplot(merged2, aes(logFC_ASD, logFC_SCZ, color = category)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +

  geom_point(alpha = 0.25, size = 1.2) +

  geom_point(
    data = lab_df,
    aes(logFC_ASD, logFC_SCZ),
    size = 2.2,
    stroke = 0.6,
    shape = 21,
    fill = "white",
    show.legend = FALSE
  ) +

  ggrepel::geom_text_repel(
    data = lab_df,
    aes(label = gene),
    size = 3,
    max.overlaps = Inf,     
    force = 2,              
    box.padding = 0.35,
    point.padding = 0.15,
    min.segment.length = 0,
    segment.color = "grey70",
    show.legend = FALSE
  ) +

  scale_color_manual(values = pal) +
  theme_classic(base_size = 13) +
  labs(
    title = "Shared DEGs",
    x = "log2FC (ASD vs Control)",
    y = "log2FC (SCZ vs Control)",
    color = "Direction"
  )
```

#Save plot above
```{r}
ggsave(
  filename = "Shared_DEG_Plot_50.eps",
  plot = last_plot(),
  device = cairo_ps,
  width = 12,
  height = 9,
  units = "in"
)
```

#Gene of Interest Investigation
#Load DGE List from EdgeR
```{r}
dge <- readRDS("DGE_List.rds")
```

#Calculate logCPM of reads 
```{r}
logCPM <- cpm(dge, log = TRUE) 

#logCPM <- as.data.frame(logCPM) %>% 
#  rownames_to_column("gene")
```

#Plot specific genes to investigate gene specific logCPM 
```{r}
gene_plot <- function(genes) {
  # Keep only genes that exist
  genes <- intersect(genes, rownames(logCPM))
  if (length(genes) == 0) {
    stop("None of the requested genes were found in rownames(dge).")
  }
  
  # logCPM[genes, ] is genes x samples; transpose so samples are rows
  df <- as.data.frame(t(logCPM[genes, , drop = FALSE]))
  
  # Add sample + condition info from dge$samples
  df$sample    <- dge$samples$sample
  df$condition <- dge$samples$condition
  
  # Long format for ggplot: one row per sample-gene
  df_long <- tidyr::pivot_longer(
    df,
    cols      = all_of(genes),
    names_to  = "gene",
    values_to = "logCPM"
  )
  
  ggplot(df_long, aes(x = sample, y = logCPM, fill = condition)) +
    geom_col() +
    facet_wrap(~ gene, scales = "free_y") +
    theme_bw(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
      title = "Gene expression (logCPM)",
      x     = "Sample",
      y     = "log2 CPM"
    )
}
```

#Save plot above 
```{r}
#svglite("NRXN1_CPM_Plot.svg", width = 14, height = 10)
gene_plot(c("CNR1","DAGLA","MGLL","PLCB1"))
#dev.off()
```

#MA Plot 
#MA Plot to check gene expression changes
```{r}
res_ASD_all <- read.csv("ControlvASD_DEGsALL.csv")
res_SCZ_all <- read.csv("ControlvSCZ_DEGsALL.csv")

ggplot(res_ASD_all, aes(x = logCPM, y = logFC)) +
  geom_point(aes(color = FDR < 0.05), alpha = 0.5, size = 1) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed") +
  theme_bw(base_size = 12) +
  labs(
    title = "MA plot: ASD vs Control",
    x = "Average expression (logCPM)",
    y = "log2 fold change (logFC)",
    color = "FDR < 0.05"
  )

ggplot(res_SCZ_all, aes(x = logCPM, y = logFC)) +
  geom_point(aes(color = FDR < 0.05), alpha = 0.5, size = 1) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed") +
  theme_bw(base_size = 12) +
  labs(
    title = "MA plot: SCZ vs Control",
    x = "Average expression (logCPM)",
    y = "log2 fold change (logFC)",
    color = "FDR < 0.05"
  )
```